\chapter*{Acknowledgments}

First and foremost, I would like to thank my supervisors, David S.~Rosenblum and Anthony Finkelstein, for their support during the past three years. I am indebted to David for his help and patience; for providing me the freedom to pursue interesting research; and for considering with me over long Skype calls the finer points of OWL-LP integration in the context of domain-specific and probabilistic model checking. Anthony's advice on authoring compelling narratives that clarify complex research, and his encouragement and strategic guidance, have been invaluable for my development as a researcher.

I have been fortunate to visit the Nebraska Intelligent MoBile Unmanned Systems (NIMBUS) Lab at the University of Nebraska-Lincoln (UNL). I would like to thank Sebastian Elbaum, co-founder of NIMBUS, for working with me to focus the research in this thesis, and for his contribution throughout my PhD\@.

I would also like to thank Emmanuel Letier for his insightful feedback during my first and second year vivas. And I would like to thank faculty and staff at the Department of Computer Science at University College London (UCL) for providing me with the resources to develop this thesis. In particular, I would like to thank Dean Mohamedally, Donald Lawrence, Graham Roberts, Jens Krinke, John Dowell, Mark Harman, Philip Treleaven and Westley Weimer for numerous valuable conversations. Last but not least, I have had the pleasure of working with talented colleagues including Jan Grochmalicki, Michal Galas and Panagiotis Papakos at UCL, and Charlie Lucas at UNL\@.

Research for this thesis was sponsored by the European Office of Aerospace Research \& Development (EOARD), a detachment of the \href{http://www.afosr.af.mil/}{Air Force Office of Scientific Research} (AFOSR), under agreement number FA8655-10-1-3007. Any conclusions, findings, opinions and recommendations expressed in this thesis are those of the author and do not necessarily reflect the views of AFOSR or EOARD\@.
